skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-07-15 paper/paper.tex
recompile

\author[FirstAuthorLastName \textit{et~al}.]{
Stephen D. Turner\,\textsuperscript{1}, V.P.
Nagraj\,\textsuperscript{1}, Matthew
Scholz\,\textsuperscript{1}, Shakeel Jessa\,\textsuperscript{1}, Carlos
Acevedo\,\textsuperscript{1}, Jianye Ge\,\textsuperscript{2}, August
Woener\,\textsuperscript{2}, Bruce Budowle\,\textsuperscript{2}
}

\address{
\textsuperscript{1}Signature Science, LLC., Charlottesville, Virginia,
22911, USA.\\
\textsuperscript{2}University of North Texas Health Science Center, Fort
Worth, Texas, 76107, USA.\\
}

@@

Inferring familial relationships between individuals using genetic data
is a common practice in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating relatedness
between samples, including genome-wide measures such as those
implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts focusing
on benchmarking these methods \citep{ramstetter2017, vries2021} have
been aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are seldom distributed with robust
documentation, test suites, or narrative examples of usage. There is a
need in the field for a well-documented software package with a
consistent design and API that contains functions to assist with
downstream manipulation, benchmarking, and analysis of SNP-based kinship
assessment methods. Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation with \textbf{R}.

@@

The skater package has several functions for importing, parsing, and
analyzing pedigree data. Pedigrees define familial relationships in a
hierarchical structure. Many genomics tools for working with pedigrees
start with a .fam file, which is a tabular format with one row per
individual and columns for unique IDs of the mother, father, and the
family unit. The skater package contains the function
\texttt{read\_fam()} to read in a PLINK-formatted .fam file, and another
function \texttt{fam2ped()} to convert the content into a pedigree
object as a nested tibble with one row per family. All pedigree
processing from skater internally leverages a data structure from the
kinship2 package \citep{kinship2}. Further functions such as
\texttt{plot\_pedigree()} produce a multi-page PDF drawing a diagram of
the pedigree for each family, while \texttt{ped2kinpair()} produces a
pairwise list of relationships between all individuals in the data with
the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to known degrees of relatedness. When
aggregated over many relationships and inferences, this can help
benchmark performance of a particular kinship analysis method. The
skater package adapts a \texttt{confusion\_matrix()} function from
\citet{clark2021} to provide standard contingency table metrics
(e.g.~sensitivity, specificity, PPV, precision, recall, F1, etc.) with a
new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more
thoroughly in the skater package vignette (see Supplementary Material),
and may be a preferrable measure of classification accuracy when
benchmarking relationship degree estimation. Taking the reciprocal of
the target and predicted degree in a typical RMSE calculation results in
larger penalties for more egregious misclassifications (e.g.,
classifying a first-degree relative pair as second degree) than
misclassifications at more distant relationships (e.g., classifying a
fourth-degree relative pair as fifth-degree).

@@

Tools such as hap-IBD \citep{zhou2020}, and IBIS \citep{seidman2020}
detect shared IBD segments between individuals. The skater package
includes functionality to take those IBD segments, compute shared
genomic centimorgan (cM) length, and convert that shared cM to a kinship
coefficient. In addition to inferred segments, these functions can
estimate ``truth'' kinship from simulated IBD segments
\citep{caballero2019}. The \texttt{read\_ibd()} function reads pairwise
IBD segments from IBD inference tools and from simulated IBD segments.
The \texttt{read\_map()} function reads in genetic map in a standard
format which is required to translate the total centimorgans shared IBD
to a kinship coefficient using the \texttt{ibd2kin()} function.

@@

\textbf{UNT FOLKS: NEED YOUR INPUT HERE!} The following rules should be
followed:

\begin{itemize}
\tightlist
\item
  The sentence should begin: `This work was supported by \ldots{}' -
\item
  The full official funding agency name should be given, i.e.~`National
  Institutes of Health', not `NIH' (full RIN-approved list of UK funding
  agencies)
\item
  Grant numbers should be given in brackets as follows: `{[}grant number
  xxxx{]}'
\item
  Multiple grant numbers should be separated by a comma as follows:
  `{[}grant numbers xxxx, yyyy{]}'
\item
  Agencies should be separated by a semi-colon (plus `and' before the
  last funding agency)
\item
  Where individuals need to be specified for certain sources of funding
  the following text should be added after the relevant agency or grant
  number `to {[}author initials{]}'.
\end{itemize}

An example is given here: `This work was supported by the National
Institutes of Health {[}AA123456 to C.S., BB765432 to M.H.{]}; and the
Alcohol \& Education Research Council {[}hfygr667789{]}.'

Oxford Journals will deposit all NIH-funded articles in PubMed Central.
See Depositing articles in repositories -- information for authors for
details. Authors must ensure that manuscripts are clearly indicated as
NIH-funded using the guidelines above.

>>>>

\author[FirstAuthorLastName \textit{et~al}.]{
Stephen D. Turner\,\textsuperscript{1}, V.P.
Nagraj\,\textsuperscript{1}, Matthew
Scholz\,\textsuperscript{1}, Shakeel Jessa\,\textsuperscript{1}, Carlos
Acevedo\,\textsuperscript{1}, Jianye Ge\,\textsuperscript{2,3}, August
Woener\,\textsuperscript{2,3}, Bruce Budowle\,\textsuperscript{2,3}
}

\address{
\textsuperscript{1}Signature Science, LLC., Charlottesville, Virginia,
22911, USA.\\
\textsuperscript{2}Center for Human Identification, University of North
Texas Health Science Center, 3500 Camp Bowie Blvd., Fort Worth, TX,
76107, USA\\
\textsuperscript{3}Department of Microbiology, Immunology, and Genetics,
University of North Texas Health Science Center, 3500 Camp Bowie Blvd.,
Fort Worth, TX, 76107, USA\\
}

@@

Inferring familial relationships between individuals using genetic data
is a common practice in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating relatedness
between samples, including genome-wide measures, such as those
implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection, such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts focusing
on benchmarking these methods \citep{ramstetter2017, vries2021} have
been aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are seldom distributed with robust
documentation, test suites, or narrative examples of usage. There is a
need in the field for a well-documented software package with a
consistent design and API that contains functions to assist with
downstream manipulation, benchmarking, and analysis of SNP-based kinship
assessment methods. Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation with \textbf{R}.

@@

The skater package has several functions for importing, parsing, and
analyzing pedigree data. Pedigrees define familial relationships in a
hierarchical structure. Many genomics tools for working with pedigrees
start with a .fam file, which is a tabular format with one row per
individual and columns for unique IDs of the mother, father, and the
family unit. The skater package contains the function
\texttt{read\_fam()} to read in a PLINK-formatted .fam file and another
function \texttt{fam2ped()} to convert the content into a pedigree
object as a nested tibble with one row per family. All pedigree
processing from skater internally leverages a data structure from the
kinship2 package \citep{kinship2}. Further functions such as
\texttt{plot\_pedigree()} produce a multi-page PDF drawing a diagram of
the pedigree for each family, while \texttt{ped2kinpair()} produces a
pairwise list of relationships between all individuals in the data with
the expected kinship coefficients for each pair.

@@

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to known degrees of relatedness. When
aggregated over many relationships and inferences, this can help
benchmark performance of a particular kinship analysis method. The
skater package adapts a \texttt{confusion\_matrix()} function from
\citet{clark2021} to provide standard contingency table metrics
(e.g.~sensitivity, specificity, PPV, precision, recall, F1, etc.) with a
new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more
thoroughly in the skater package vignette (see Supplementary Material)
and may be a preferable measure of classification accuracy when
benchmarking relationship degree estimation. Taking the reciprocal of
the target and predicted degree in a typical RMSE calculation results in
larger penalties for more egregious misclassifications (e.g.,
classifying a first-degree relative pair as second-degree) than
misclassifications at more distant relationships (e.g., classifying a
fourth-degree relative pair as fifth-degree).

@@

Tools such as hap-IBD \citep{zhou2020}, and IBIS \citep{seidman2020}
detect shared IBD segments between individuals. The skater package
includes functionality to take those IBD segments, compute shared
genomic centimorgan (cM) length, and converts that shared cM to a
kinship coefficient. In addition to inferred segments, these functions
can estimate ``truth'' kinship from simulated IBD segments
\citep{caballero2019}. The \texttt{read\_ibd()} function reads pairwise
IBD segments from IBD inference tools and from simulated IBD segments.
The \texttt{read\_map()} function reads in genetic map in a standard
format which is required to translate the total centimorgans shared IBD
to a kinship coefficient using the \texttt{ibd2kin()} function.

@@

This work was supported in part by award 2019-DU-BX-0046 (Dense DNA Data
for Enhanced Missing Persons Identification) to B.B., awarded by the
National Institute of Justice, Office of Justice Programs, U.S.
Department of Justice and by internal funds from the Center for Human
Identification. The opinions, findings, and conclusions or
recommendations expressed are those of the authors and do not
necessarily reflect those of the U.S. Department of Justice.
